Capturing the geometric structure of our experiences and how we remember them

### 2018-08-16 paper/main.tex
trimmed list-learning section and moved the shape analyses up

First, we asked whether the estimated number of recall events ($k$) by participant was related to hand-annotated accuracy as published in Chen et al. (2017).  We found a strong positive correlation where subjects with a greater number of recall events also had better overall memory performance (Pearson's $r$(16)=.67, $p$=.003). Then, we considered how participants initiated the recall sequence (known in the literature as the `probability of first recall' or `PFR'). This metric quantifies the likelihood of initiating a sequence of recalls based on an item's serial position during encoding. We found that participants tended to initiate their recall sequences with the first few events (Fig.~\ref{fig:list-learning}a), which is qualitatively very similar to previously published list learning experiments (REF?). Next, we considered another well-studied memory measure in the list-learning literature, the lag conditional response probability curve (or lag-CRP) ~\citep{Kaha96}. Given the recall of a stimulus in study position $n$, the lag-CRP quantifies the probability that the next item recalled will be of lag $i$. The result suggests a strong bias to transition sequentially events in the forward direction (Fig.~\ref{fig:list-learning}b). This can be seen by the high probability values in the lag ~1-5 positions compared to the rest of the transition possibilties. Notably, participants were instructed to recall as much as they could in order, and the shape of this lag-CRP is qualitatively similar to list-learning studies of serial recall (REF?), where participants are required to recall lists of words in their serial order. Finally, we assessed memory performance for each event in the video as a function of its serial position during encoding (Fig.~\ref{fig:list-learning}c). We found that there was substantial variability in memory for the different events (STAT). Furthermore, the shape of the curve is different than what is typically found in list learning experiments.

 We also considered two additional across-participant measures of recall that characterize memory organization: temporal clustering and semantic clustering. Temporal clustering (REF?) measures the extent to which participants tend to cluster recalls of events encoded nearby in time.  For example, if a participant recalled in exact serial order, the score would be 1 whereas if they sequentially recalled randomly positioned items the score would be \~.5.  We found that participants who clustered in time also recalled a greater number of events (Pearson's $r$(16)=.62, $p$=.007). Next, we assessed semantic clustering.  Similar to temporal clustering, semantic clustering measures the extent to which participants cluster their responses by semantic similarity.  We used the topic vector representations of each event as a proxy for semantic information contained in that event.  For each transition, we computed the rank percentile of the of the correlation between the the current event and then next event. For example, if a participant sequentially recalled events where each successive recall was the event that was most semantically similar to the previous, the semantic clustering score would be 1.  If the events were recalled in a random order with respect to semantic information, the clustering score would be ~.5. We found that the semantic clustering score was related to memory performance across participants (Pearson's $r$(16)=.55, $p$=.02).  Thus, participants who organized their recalls with respect to the semantic information contained in the scene had better memory performance.

@@

Studying episodic memory is commonly distilled down to a process of matching specific moments of a past experience with specific mnemonic outcomes.  In traditional trial-based free recall and item recognition experiments, individual stimuli encountered during encoding are typically labeled as ``remembered'' or ``forgotten'' depending on whether the stimulus is recalled/recognized during a subsequent test. However, this approach does not translate well to naturalistic experiences. For one thing, the contents of our recall at any given moment might be reflected in many prior experiences/moments. Furthermore, the particular words used to describe the experience will inevitably vary across people and even across repeated recollections within an individual. Thus, there is not a ``one-to-one'' mapping between naturalistic experiences and their mnemonic counterparts. Our topic modeling approach, whereby we consider the broad ``theme'' present in different moments of participants' experiences and their memories for those experiences, allows us to flexibly and accurately characterize memory for naturalistic experiences. This allows us to quantify which moments from the past the current thoughts match, in terms of their thematic content and critically, our ability to perform this matching does not require participants to use any specific overlapping words.

Historically, human memory reseachers have focussed on measuring the quantity of information recovered by an individual under different experimental conditions (i.e. the proportion correct). However, for real world experiences this is not a well-defined task. For example, remembering the patterns and colors of each person's shirt in a crowd might be considered as excellent recall in a standard memory task setup. But if the rememberer failed to note that the people in the crowd were gathered for their surprise birthday party, then they would have missed the "point" of the experience. Our work characterizes experience by comparing the overall ``shape'' of an experience and a memory. We assess the quality of memory for the movie participants viewed by measuring the match between the shapes of the movie's trajectory and each participant's recall trajectory. By contrast, the number of recalls could be captured by the "sampling frequency" along that trajectory-- but the number of recalls alone cannot tell us whether participants successfully recollected the meaning of the story by capturing the salient points of the narrative that define its main shape.

>>>>

First, we asked whether the estimated number of recall events ($k$) by participant was related to hand-annotated accuracy as published in Chen et al. (2017).  We found a strong positive correlation where subjects with a greater number of recall events also had better overall memory performance (Pearson's $r$(16)=.67, $p$=.003). Then, we considered how participants initiated the recall sequence (known in the literature as the `probability of first recall' or `PFR'). We found that participants tended to initiate their recall sequences with the first few events (Fig.~\ref{fig:list-learning}a), which is qualitatively very similar to previously published list learning experiments (REF?). Next, we considered another well-studied memory measure in the list-learning literature, the lag conditional response probability curve (or lag-CRP) ~\citep{Kaha96}. The result suggests a strong bias to transition sequentially events in the forward direction (Fig.~\ref{fig:list-learning}b). Finally, we assessed memory performance for each event in the video as a function of its serial position during encoding (Fig.~\ref{fig:list-learning}c). We found that there was substantial variability in memory for the different events (STAT).

We also considered two additional across-participant measures of recall that characterize memory organization: temporal clustering and semantic clustering. We found that participants who clustered in time also recalled a greater number of events (Pearson's $r$(16)=.62, $p$=.007). Next, we assessed semantic clustering. We found that the semantic clustering score was related to memory performance across participants (Pearson's $r$(16)=.55, $p$=.02).  Thus, participants who organized their recalls with respect to the semantic information contained in the scene had better memory performance.

@@

Studying episodic memory is commonly distilled down to a process of matching specific moments of a past experience with specific mnemonic outcomes.  In traditional trial-based free recall and item recognition experiments, individual stimuli encountered during encoding are typically labeled as ``remembered'' or ``forgotten'' depending on whether the stimulus is recalled/recognized during a subsequent test. While this approach has advanced our understanding of human memory immensely, it does not translate well to naturalistic experiences. For one thing, the contents of our recall at any given moment might be reflected in many prior experiences/moments. Furthermore, the particular words used to describe the experience will inevitably vary across people and even across repeated recollections within an individual. Thus, there is not a ``one-to-one'' mapping between naturalistic experiences and their mnemonic counterparts. Our topic modeling approach, whereby we consider the broad ``theme'' present in different moments of participants' experiences and their memories for those experiences, affords us the ability to flexibly and accurately characterize memory for naturalistic experiences. This approach allows us to quantify which moments from the past and the current recounted experience match in terms of their thematic content and critically, our ability to perform this matching does not require participants to use any specific overlapping words.

Historically, human memory reseachers have focussed on measuring the quantity of information recovered by an individual (i.e. the proportion correct) under different experimental conditions. However, for real world experiences this is not a well-defined task. For example, remembering the patterns and colors of each person's shirt in a crowd might be considered as excellent recall in a standard memory task setup. But if the rememberer failed to note that the people in the crowd were gathered for their surprise birthday party, then they would have missed the "point" of the experience. Our work characterizes experience by comparing the overall ``shape'' of a dynamic stimulus and a memory. We assess the quality of memory for the movie participants viewed by measuring the match between the shapes of the movie's trajectory and each participant's recall trajectory. By contrast, the number of recalls could be captured by the "sampling frequency" along that trajectory-- but the number of recalls alone cannot tell us whether participants successfully recollected the meaning of the story by capturing the salient points of the narrative that define its main shape.
